{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-23 thesis/quicksand.tex
Data-race (haha)
やはり data race analysis

\qrevisionminor{How should a model checker know to instrument this particular data race for preemption
in order to find the assertion failure lurking underneath?}
Committing in advance to
preempt on every instruction is certain to include this data race,
but invites massive state space explosion.
Even as DPOR helps to skip equivalent interleavings of non-conflicting transitions,
DPOR itself is $O(n^2)$ in the number of preemption points in a single execution,
which is not compatible with such an approach.
Accordingly, stateless model checkers must find more efficient ways to be able to uncover bugs such as these.
%
In related work,
Portend~\cite{portend} proposed to combine data race analysis with preemption-driven artificial scheduling,
although it obtains its data-race candidates from a stand-alone, single-pass analysis.
In order to
%make sure we
identify every
%one of a program's
data race that could possibly arise under the given test case,
a model checker must check many different interleavings to begin with,
perform the Portend approach for every data race it finds,
which may in turn uncover more data races
(hidden in flow control paths reachable only through interleavings of the first race, perhaps),
and then continue model checking those multiple races together
in a
%sort of
bidirectional feedback loop between the two algorithms.
Upcoming,
\cref{sec:quicksand-id} will show how I achieve this in Quicksand, and
\cref{sec:quicksand-soundness} will justify the technique's formal verification power.
In the evaluation, %(\cref{sec:quicksand-eval}),
\qrevisionminor{\cref{sec:quicksand-eval-bugs} and \cref{sec:quicksand-eval-verif}
will show that Quicksand strikes a healthy balance between fast bug-finding and full verification,
and
\cref{sec:quicksand-eval-nondets}
%in particular
will justify the need
for such a feedback loop %between model checking and data race analysis.
by showing that many
data races require model checking to reliably detect.
%bugs require nondeterministic data races to expose.
}

@@

\begin{itemize}
	\item {\bf Single-state-space model checking} refers to the state-of-the-art model checking strategy,
		i.e., approaching each test with preemption points fixed in advance.
	\item {\bf Maximal state space} refers to the set of thread interleavings possible
		by preempting on all currently-known preemption points,
		whether synchronization or data-race,
		i.e., the singular state space tested by {\em single-state-space model checking}.
	\item {\bf Minimal state space} indicates the opposite:
		those thread interleavings requiring no more than Landslide's mandatory preemptions on voluntary context switches.
	\item {\bf Data-race bug}
		will refer to a concrete failure, such as \Cref{fig:pps-example}'s assertion failure above, whereas
	\item {\bf Data race} refers to the racing access pair itself (\cref{sec:landslide-datarace}).
		% TODO: scan this chapter for consistency of dis
	\item {\bf Data-race candidate} shall refer specifically to
		potentially-racing accesses identified by Limited Happens-Before, %(\cref{sec:landslide-lhb}), % sub sub sec
		when disambiguation with {\em data races} is necessary.
	\item {\bf Benign data races} are those that, when reordered in an alternate interleaving,
		do not lead to a {\em data-race bug}, while
	\item {\bf False positives} are {\em data-race candidates} that,
		upon trying to reorder them,
		turn out not to exist in the alternate interleaving at all,
		such as in \Cref{fig:hb-example}(b).
	\item {\bf Nondeterministic data race}
		will refer to data races that cannot be exposed on the first thread interleaving,
		but require model checking to expose to begin with.
\end{itemize}
}

@@

Given unlimited CPU time for testing,
one would always want to switch to the new maximal state space whenever adding a new preemption point.
The maximal state space is guaranteed to subsume all execution sequences reachable in any subset state space,
so considering any incomplete subset of the known preemption points would be redundant work.
\Cref{alg:algorithm0} demonstrates this na\"ive approach.
It is seeded with the set of all statically-known synchronization API preemption points,
and invoked whenever a new data-race candidate is found.
%
The upcoming proofs in \cref{sec:quicksand-soundness},
being concerned with the verification guarantee provided when the search may complete within the CPU budget,
are based on this simple version of Iterative Deepening.
The user may also wish to configure her testing tool to prefer this approach, at her discretion,
such as when she believes all bugs have been fixed and wants a verification as fast as possible;
% TODO: clarify sec reference to -M flag
\cref{sec:quicksand-implementation} discusses this execution mode further.

@@

The new state spaces may expose a failure, in which case Iterative Deepening must stop and report a data-race bug,
or complete successfully, indicating a {\em benign} (i.e., false-positive) data race.
They may also uncover a new data-race candidate entirely in some alternate interleaving,
in which case we may iteratively advance to a superset state space which will preempt at both racing access pairs.
Being constrained by a CPU budget,
Iterative Deepening may time out before completing a data race's associated state space,
in which case the data race remains neither confirmed nor refuted.
%Depending on how much burden the implementation wants to impose on the user,
%it may then report it as a
%report a potential false positive that the user must handle
In such cases, Quicksand elects to impose some burden on the user
by reporting it as a potential false positive
and recommend that she investigate it by hand to judge for herself whether it be a bug.
\cref{sec:quicksand-discussion-partial} will discuss future opportunities for improving
debugging output in cases of such {\em partial verification}.
However, experience shows that this interactivity pays off:
in the next chapter's educational user study (\cref{sec:education-eval}),
one student reported during the survey that they used this recommendation,
combined with their own intuition,
to find a bug that Quicksand was not able to find alone (\cref{sec:education-reasons-worthwhile}).

@@

Landslide classifies data-race candidates as {\em both-order} or {\em single-order},
as defined in prior work \cite{portend},
based on whether it observed the racing instructions ordered in both possible sequences or only one
in the original state space, respectively.
Single-order candidates are more likely to be false positives (\cref{sec:background-datarace}),
although preempting during the access itself is necessary to say for sure.
Hence, Quicskand add preemption points for both types of candidates,
and heuristically prioritizes jobs with both-order data races
over those with only single-order data races.
The HeuristicPriority($\alpha$) call in \Cref{alg:handledatarace} corresponds to this strategy.
For single-order races, Quicksand does not initially add a preemption point for the later access at all:
if preempting on the first access is capable of reordering the race,
it will be updated to both-order in the new state space, and the second preemption point will be added then.
\cref{sec:warpzone-heuristics} will discuss opportunities for future work to expand
these heuristics with more nuanced search strategies still.

@@

Finally, I identified a particular class of false positive data-race candidates
under the Limited Happens-Before analysis (\cref{sec:background-datarace})
in which the associated memory was recycled by re-allocation between the two accesses,
and claim that it is safe to completely disregard them when considering where to add new preemption points.
Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behaviour.
If the {\tt malloc} on line~4 returns the same address passed to {\tt free} on line~2,
then lines~1 and 7 will be flagged as a potential data race.
I term this a {\em reallocation false-positive data-race candidiate}.
To the human eye, this is obviously a false positive:
reordering lines~4-7 before lines~1-2 will cause {\tt malloc} to return a different region of allocated memory,
in turn causing {\tt x} and {\tt y} to no longer collide.
In studying a similar pattern, the Eraser tool from prior work \cite{eraser}
found that Thread 2's logic usually corresponds to an initialization pattern,
but for generality I have added an arbitrary {\tt publish} action to the example on line~6.

@@

As long as the allocation heap is properly synchronized,
a Pure Happens-Before analysis should identify a happens-before edge
between line~2's {\tt free()} and line~4's {\tt malloc()},
and report no race.
However, the upcoming evaluation will show that Limited Happens-Before retains some advantages over Pure
(\cref{sec:quicksand-eval-bugs}),
so it is useful to
%be able to
automatically suppress data-race candidates
that are certain to end up being false positives when reordered.
Such collisions could instead be avoided with a hacked allocator which never recycles memory,
but that could unacceptably impact performance in {\tt malloc()}-heavy tests.

@@

{\bf Assumptions.}
The proofs are built on a DPOR definition which assumes sequentially-consistent memory hardware.
All algorithms involved are assumed to operate on a machine model of a single globally-consistent execution trace,
which fundamentally cannot account for memory reordering nondeterminism.
%\cref{sec:quicksand-discussion} discusses this limitation further; % that's a lie..
For existing work on combining DPOR with relaxed memory, I refer the reader to \cite{tsopso}.
They also assume the Limited Happens-Before definition
for the data-race analysis.
I leave the case for Pure Happens-Before to future work,
although if I may appeal to intuition,
it requires only to show that for any data-race candidate
Limited Happens-Before reports in a given execution,
that Pure Happens-Before does not,
either it will be a false positive,
or the latter will find it in an alternate execution within the same state space,
or the latter will find a different data race that ultimately leads to a bigger state space in which the first one may be found,
much like a generalization of \cref{sec:quicksand-realloc}.

@@

\begin{theorem}[Soundness of eliminating reallocation data-race candidates]
        If a reallocation candidate is not a false positive,
DPOR will reorder threads such that
%DPOR will test an alternate thread interleaving in which
either
the accesses can race without fitting the reallocation pattern,
or a use-after-free bug will be reported immediately.
\end{theorem}

@@

The interface with the model checker has two parts.
First, when starting each job, Quicksand creates a configuration file declaring which preemption points to use,
plus other test-case-specific options such as
which preemption points to suppress (\cref{sec:landslide-pps}),
especially those arising from the {\tt malloc} lock (\cref{sec:quicksand-realloc}),
which functions DPOR and data-race analysis should treat as trusted code (\cref{sec:landslide-config-landslide}),
whether to enable mutex-testing mode (\cref{sec:landslide-dynamicconfig}),
transactional memory (\cref{sec:tm-implementation}),
and so on.
This is done by {\tt run\_job()} in {\tt job.c}.

Then, a dedicated Quicksand thread
({\tt start\_job()} in {\tt job.c})
communicates with the MC process via message-passing
over a FIFO pipe % not gonna cite io.c c.c
({\tt talk\_to\_child()} in {\tt messaging.c}).
Landslide messages after testing each interleaving to report updated progress and ETAs
as well as whenever a new data-race candidate or bug is found.
Quicksand in turn replies whether to resume/suspend (due to too high ETA) or quit (due to timeout)
({\tt handle\_should\_continue()} in {\tt messaging.c}).
It suspends jobs simply by making Landslide wait on a message-passing reply.
Should Quicksand later re-schedule a suspended job, it sends a message to continue,
resuming Landslide right where it left off;
otherwise, it wakes it up only after time runs out, causing it to exit immediately.
The message-passing format is defined at the top of {\tt messaging.c},
and a matching definition appears in Landslide's source file of the same name.

@@

The other end of the messaging protocol (\cref{sec:quicksand-impl-mc})
is implemented in {\tt messaging.c}.
When Quicksand suspends Landslide,
it detects how much time it spent asleep,
and corrects for that amount during its next ETA computation ({\tt fudge\_time()} in {\tt estimate.c}).
Landslide's data race analysis also includes a heuristic to avoid reporting ``too suspicious''
data-race candidates which it believes arise from the initialization pattern \cite{eraser}:
if a conflicting access pair is single-order (\cref{sec:quicksand-heuristics})
and also arose during a known synchronization API's {\tt init()} or {\tt destroy()} function,
Landslide will not message it to Quicksand,
at least not until it is reclassified as both-order.

@@

The QS-Sync-Only experiment tests whether Iterative Deepening would be effective
even for model checking domains without data races,
such as distributed systems \cite{macemc,modist,samc,dbug-retreat,concuerror}
and programming languages whose type systems statically reject concurrent mutable shared state
\cite{erlang,haskell,rust-book}.
When Quicksand ignores all data-race candidates,
its results are competitive with SSS-MC-DPOR, although SSS-MC-ICB outperforms it slightly.
This is unsurprising: the seed subsets of preemption points
that QS-Sync-Only is limited to
(\cref{sec:quicksand-initial-pps})
are much less flexible than ICB's preemption strategy.
This result suggests that in future work,
Quicksand should consider using ICB in parallel with its default configuration when it finds no data-race candidates to test.
I discuss this possibility further in \cref{sec:warpzone-heuristics}.

@@

Ultimately, using Limited Happens-Before for finding data-race candidates allowed Quicksand to find more bugs,
while Pure Happens-Before allowed for reaching full verification faster.
I attribute this trade-off to the fact that
Limited Happens-Before need not wait to test many alternate thread interleavings before
finding a data-race candidate to begin with;
%a potential data-race candidate is confirmed;
rather, it can add new jobs to start testing potential races immediately.%
\footnote{Upcoming, \Cref{tab:drstatistix} will corroborate this conclusion:
the difference is most dramatic in {\tt alarm-\allowbreak{}simultaneous},
the test where Quicksand struggled most to finish even small subset jobs.}
%
On the other hand, Limited Happens-Before can get overwhelmed by too many false positives,
needing to refute such candidates by testing new state spaces for each one,
while Pure Happens-Before can refute false positives {\em en passant}
by testing alternate interleavings in its original state spaces.
%In both cases, the performance differences seem to become significant only after about 10 minutes of CPU time,
%which suggests... I don't actually know!
This suggests that
%each approach has merit, and that
model checkers which incorporate data-race analysis should implement both modes
and offer the user to choose based on their desired and/or expected testing outcome.

@@

Beyond finding new bugs and completing full verifications with data-race preemption points,
I evaluated Quicksand's performance for classifying data-race candidates in two ways:
its ability to check nondeterministic data races not reachable under a single-pass analysis (\cref{sec:quicksand-pps})
and its ability to suppress reallocation false positives (\cref{sec:quicksand-id-realloc}).
\Cref{tab:drstatistix} presents the results for this section.

@@

Some memory accesses may be hidden in a control flow path that requires a nondeterministic preemption to be executed
(\cref{sec:quicksand-pps}).
In such cases, a single-pass dynamic data-race detector
might not achieve the coverage necessary
to identify a racing access pair as a candidate at all,
let alone check the resulting behaviour with such as Landslide.
I instrumented Landslide to report these to Quicksand
and counted how many such led to Quicksand finding new bugs when used as preemption points.
Such bugs could be considered {\em false negatives} of the single-pass approach.
The left half of \Cref{tab:drstatistix}
breaks down the types of bugs
found in each test case,
showing both the total number of data-race bugs
and the number among those that required such nondeterministic data races to expose.
% I denote these data race reports as {\em nondeterministic},
% while those that could be found on the first interleaving .....
%
To ensure a fair comparison, I disabled Quicksand's reallocation false positive suppression
(\cref{sec:quicksand-id-realloc}, itself evaluated in the next section)
for this experiment.
This prevents Landslide from suppressing an observed reallocation data-race candidate on the first interleaving,
which would falsely classify it as nondeterministic,
even though a single-pass would not (indeed, should not) suppress such candidates.

\Cref{fig:dr-falsenegs}
%compares the types of data-race candidates necessary to expose each data-race bug in the test suite.
visualizes the difference between single-pass and model-checking-enabled data race analysis.
The first and third series represent the bugs found using preemption points from single-pass data-race candidates only,
% not entirely true, as portend could be given data-race traces from an MC
%but they don't do it in their paper, so i feel comfortable making this claim
i.e., the state-of-the-art approach used by RaceFuzzer \cite{racefuzzer} and Portend \cite{portend}.
The second and fourth series show all data-race bugs Quicksand found,
which includes the former type as well as new bugs involving nondeterministic races.
QS-Limited-HB found a nice 69 data-race bugs in total, % nice
15 of which %could not be found with single-pass data-race candidates alone.
required nondeterministic data-race preemption points to expose.
QS-Pure-HB is even more dependent thereupon, %on nondeterministic data-race preemption points,
%requiring nondeterministic data-race preemption points
requiring them in 21 cases among its 44 total data-race bugs.
Moreover, although the frequency
of these nondeterministic races varies across the different test cases
(for example, almost all in {\tt broadcast\_test} were nondeterministic; almost none in {\tt mutex\_test}),
they are still at least present in all tests,
meaning it is not just an issue of writing ``better'' test cases to avoid them.

@@

Note that I am not comparing how much testing time is required before identifying the data-race candidates involved in each bug.
Single-pass data races can all be found after a single program execution,
while Quicksand may potentially take up to all 10 CPU-hours before identifying a nondeterministic data race.
However, prior work data-race tools \cite{tsan,fasttrack},
being not integrated with a model checker,
are not intended to discover new candidates under subsequent runs.
Running a single-pass data-race tool repeatedly for 10 CPU-hours could potentially uncover some nondeterministic candidates,
but stress testing's comparative problem with achieving reliable coverage is already well-understood
\cite{chess-icb,gambit},
so I hope the reader will consider this experiment enough evidence for model checking already.
Likewise, replay-based tools such as RaceFuzzer \cite{racefuzzer} and Portend \cite{portend}
depend upon the data-race detector to provide an execution trace leading to each candidate.
This result suggests that
such tools could benefit from a similar feedback loop as is used in Iterative Deepening,
for example, to discover new transitively-reachable data races while testing initial ones,
even if full verification not necessarily be their goal.

@@

In \cref{sec:quicksand-realloc} I showed the soundness of
suppressing data race reports between two heap accesses when the surrounding memory was re-allocated in between.
\Cref{tab:drstatistix}'s ``Realloc FPs'' column shows the total number of such data-race candidates for each test program,
totaling 965 across all tests.
% NB. Calculated this by grepping QS-LHB id-logs for "for realsies" (see quicksand pp.c).
Among these, only 64 were observed to avoid the reallocation in an alternate interleaving,
thereupon being promoted to real data-race preemption points.
\cref{sec:quicksand-realloc}'s proof guarantees the safety of pruning all state spaces resulting from the 901 others.
%
Among the 64 true data-races, %which initially fit the malloc-recycle pattern,
none exposed a new bug when used as a preemption point.
This suggests that for other data-race tools,
suppressing reallocation candidates may be a productive heuristic,
even if unsound without Iterative Deepening.
However, Quicksand was able to correctly identify the 64 violations of that heuristic (among 26 distinct tests),
and fall back to classifying them with DPOR.

>>>>

\qrevisionminor{How should a model checker know to instrument this particular data race for preemption
in order to find the assertion failure lurking underneath?}
Committing in advance to
preempt on every instruction is certain to include this data race,
but invites massive state space explosion.
Even as DPOR helps to skip equivalent interleavings of non-conflicting transitions,
DPOR itself is $O(n^2)$ in the number of preemption points in a single execution,
which is not compatible with such an approach.
Accordingly, stateless model checkers must find more efficient ways to be able to uncover bugs such as these.
%
In related work,
Portend~\cite{portend} proposed to combine data race analysis with preemption-driven artificial scheduling,
although it obtains its data race candidates from a stand-alone, single-pass analysis.
In order to
%make sure we
identify every
%one of a program's
data race that could possibly arise under the given test case,
a model checker must check many different interleavings to begin with,
perform the Portend approach for every data race it finds,
which may in turn uncover more data races
(hidden in flow control paths reachable only through interleavings of the first race, perhaps),
and then continue model checking those multiple races together
in a
%sort of
bidirectional feedback loop between the two algorithms.
Upcoming,
\cref{sec:quicksand-id} will show how I achieve this in Quicksand, and
\cref{sec:quicksand-soundness} will justify the technique's formal verification power.
In the evaluation, %(\cref{sec:quicksand-eval}),
\qrevisionminor{\cref{sec:quicksand-eval-bugs} and \cref{sec:quicksand-eval-verif}
will show that Quicksand strikes a healthy balance between fast bug-finding and full verification,
and
\cref{sec:quicksand-eval-nondets}
%in particular
will justify the need
for such a feedback loop %between model checking and data race analysis.
by showing that many
data races require model checking to reliably detect.
%bugs require nondeterministic data races to expose.
}

@@

\begin{itemize}
	\item {\bf Single-state-space model checking} refers to the state-of-the-art model checking strategy,
		i.e., approaching each test with preemption points fixed in advance.
	\item {\bf Maximal state space} refers to the set of thread interleavings possible
		by preempting on all currently-known preemption points,
		whether synchronization or data-race,
		i.e., the singular state space tested by {\em single-state-space model checking}.
	\item {\bf Minimal state space} indicates the opposite:
		those thread interleavings requiring no more than Landslide's mandatory preemptions on voluntary context switches.
	\item {\bf Data-race bug}
		will refer to a concrete failure, such as \Cref{fig:pps-example}'s assertion failure above, whereas
	\item {\bf Data race} refers to the racing access pair itself (\cref{sec:landslide-datarace}).
		% TODO: scan this chapter for consistency of dis
	\item {\bf Data race candidate} shall refer specifically to
		potentially-racing accesses identified by Limited Happens-Before, %(\cref{sec:landslide-lhb}), % sub sub sec
		when disambiguation with {\em data races} is necessary.
	\item {\bf Benign data races} are those that, when reordered in an alternate interleaving,
		do not lead to a {\em data-race bug}, while
	\item {\bf False positives} are {\em data race candidates} that,
		upon trying to reorder them,
		turn out not to exist in the alternate interleaving at all,
		such as in \Cref{fig:hb-example}(b).
	\item {\bf Nondeterministic data race}
		will refer to data races that cannot be exposed on the first thread interleaving,
		but require model checking to expose to begin with.
\end{itemize}
}

@@

Given unlimited CPU time for testing,
one would always want to switch to the new maximal state space whenever adding a new preemption point.
The maximal state space is guaranteed to subsume all execution sequences reachable in any subset state space,
so considering any incomplete subset of the known preemption points would be redundant work.
\Cref{alg:algorithm0} demonstrates this na\"ive approach.
It is seeded with the set of all statically-known synchronization API preemption points,
and invoked whenever a new data race candidate is found.
%
The upcoming proofs in \cref{sec:quicksand-soundness},
being concerned with the verification guarantee provided when the search may complete within the CPU budget,
are based on this simple version of Iterative Deepening.
The user may also wish to configure her testing tool to prefer this approach, at her discretion,
such as when she believes all bugs have been fixed and wants a verification as fast as possible;
% TODO: clarify sec reference to -M flag
\cref{sec:quicksand-implementation} discusses this execution mode further.

@@

The new state spaces may expose a failure, in which case Iterative Deepening must stop and report a data-race bug,
or complete successfully, indicating a {\em benign} (i.e., false-positive) data race.
They may also uncover a new data race candidate entirely in some alternate interleaving,
in which case we may iteratively advance to a superset state space which will preempt at both racing access pairs.
Being constrained by a CPU budget,
Iterative Deepening may time out before completing a data race's associated state space,
in which case the data race remains neither confirmed nor refuted.
%Depending on how much burden the implementation wants to impose on the user,
%it may then report it as a
%report a potential false positive that the user must handle
In such cases, Quicksand elects to impose some burden on the user
by reporting it as a potential false positive
and recommend that she investigate it by hand to judge for herself whether it be a bug.
\cref{sec:quicksand-discussion-partial} will discuss future opportunities for improving
debugging output in cases of such {\em partial verification}.
However, experience shows that this interactivity pays off:
in the next chapter's educational user study (\cref{sec:education-eval}),
one student reported during the survey that they used this recommendation,
combined with their own intuition,
to find a bug that Quicksand was not able to find alone (\cref{sec:education-reasons-worthwhile}).

@@

Landslide classifies data race candidates as {\em both-order} or {\em single-order},
as defined in prior work \cite{portend},
based on whether it observed the racing instructions ordered in both possible sequences or only one
in the original state space, respectively.
Single-order candidates are more likely to be false positives (\cref{sec:background-datarace}),
although preempting during the access itself is necessary to say for sure.
Hence, Quicskand add preemption points for both types of candidates,
and heuristically prioritizes jobs with both-order data races
over those with only single-order data races.
The HeuristicPriority($\alpha$) call in \Cref{alg:handledatarace} corresponds to this strategy.
For single-order races, Quicksand does not initially add a preemption point for the later access at all:
if preempting on the first access is capable of reordering the race,
it will be updated to both-order in the new state space, and the second preemption point will be added then.
\cref{sec:warpzone-heuristics} will discuss opportunities for future work to expand
these heuristics with more nuanced search strategies still.

@@

Finally, I identified a particular class of false positive data race candidates
under the Limited Happens-Before analysis (\cref{sec:background-datarace})
in which the associated memory was recycled by re-allocation between the two accesses,
and claim that it is safe to completely disregard them when considering where to add new preemption points.
Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behaviour.
If the {\tt malloc} on line~4 returns the same address passed to {\tt free} on line~2,
then lines~1 and 7 will be flagged as a potential data race.
I term this a {\em reallocation false-positive data race candidiate}.
To the human eye, this is obviously a false positive:
reordering lines~4-7 before lines~1-2 will cause {\tt malloc} to return a different region of allocated memory,
in turn causing {\tt x} and {\tt y} to no longer collide.
In studying a similar pattern, the Eraser tool from prior work \cite{eraser}
found that Thread 2's logic usually corresponds to an initialization pattern,
but for generality I have added an arbitrary {\tt publish} action to the example on line~6.

@@

As long as the allocation heap is properly synchronized,
a Pure Happens-Before analysis should identify a happens-before edge
between line~2's {\tt free()} and line~4's {\tt malloc()},
and report no race.
However, the upcoming evaluation will show that Limited Happens-Before retains some advantages over Pure
(\cref{sec:quicksand-eval-bugs}),
so it is useful to
%be able to
automatically suppress data race candidates
that are certain to end up being false positives when reordered.
Such collisions could instead be avoided with a hacked allocator which never recycles memory,
but that could unacceptably impact performance in {\tt malloc()}-heavy tests.

@@

{\bf Assumptions.}
The proofs are built on a DPOR definition which assumes sequentially-consistent memory hardware.
All algorithms involved are assumed to operate on a machine model of a single globally-consistent execution trace,
which fundamentally cannot account for memory reordering nondeterminism.
%\cref{sec:quicksand-discussion} discusses this limitation further; % that's a lie..
For existing work on combining DPOR with relaxed memory, I refer the reader to \cite{tsopso}.
They also assume the Limited Happens-Before definition
for the data race analysis.
I leave the case for Pure Happens-Before to future work,
although if I may appeal to intuition,
it requires only to show that for any data race candidate
Limited Happens-Before reports in a given execution,
that Pure Happens-Before does not,
either it will be a false positive,
or the latter will find it in an alternate execution within the same state space,
or the latter will find a different data race that ultimately leads to a bigger state space in which the first one may be found,
much like a generalization of \cref{sec:quicksand-realloc}.

@@

\begin{theorem}[Soundness of eliminating reallocation data race candidates]
        If a reallocation candidate is not a false positive,
DPOR will reorder threads such that
%DPOR will test an alternate thread interleaving in which
either
the accesses can race without fitting the reallocation pattern,
or a use-after-free bug will be reported immediately.
\end{theorem}

@@

The interface with the model checker has two parts.
First, when starting each job, Quicksand creates a configuration file declaring which preemption points to use,
plus other test-case-specific options such as
which preemption points to suppress (\cref{sec:landslide-pps}),
especially those arising from the {\tt malloc} lock (\cref{sec:quicksand-realloc}),
which functions DPOR and data race analysis should treat as trusted code (\cref{sec:landslide-config-landslide}),
whether to enable mutex-testing mode (\cref{sec:landslide-dynamicconfig}),
transactional memory (\cref{sec:tm-implementation}),
and so on.
This is done by {\tt run\_job()} in {\tt job.c}.

Then, a dedicated Quicksand thread
({\tt start\_job()} in {\tt job.c})
communicates with the MC process via message-passing
over a FIFO pipe % not gonna cite io.c c.c
({\tt talk\_to\_child()} in {\tt messaging.c}).
Landslide messages after testing each interleaving to report updated progress and ETAs
as well as whenever a new data race candidate or bug is found.
Quicksand in turn replies whether to resume/suspend (due to too high ETA) or quit (due to timeout)
({\tt handle\_should\_continue()} in {\tt messaging.c}).
It suspends jobs simply by making Landslide wait on a message-passing reply.
Should Quicksand later re-schedule a suspended job, it sends a message to continue,
resuming Landslide right where it left off;
otherwise, it wakes it up only after time runs out, causing it to exit immediately.
The message-passing format is defined at the top of {\tt messaging.c},
and a matching definition appears in Landslide's source file of the same name.

@@

The other end of the messaging protocol (\cref{sec:quicksand-impl-mc})
is implemented in {\tt messaging.c}.
When Quicksand suspends Landslide,
it detects how much time it spent asleep,
and corrects for that amount during its next ETA computation ({\tt fudge\_time()} in {\tt estimate.c}).
Landslide's data race analysis also includes a heuristic to avoid reporting ``too suspicious''
data race candidates which it believes arise from the initialization pattern \cite{eraser}:
if a conflicting access pair is single-order (\cref{sec:quicksand-heuristics})
and also arose during a known synchronization API's {\tt init()} or {\tt destroy()} function,
Landslide will not message it to Quicksand,
at least not until it is reclassified as both-order.

@@

The QS-Sync-Only experiment tests whether Iterative Deepening would be effective
even for model checking domains without data races,
such as distributed systems \cite{macemc,modist,samc,dbug-retreat,concuerror}
and programming languages whose type systems statically reject concurrent mutable shared state
\cite{erlang,haskell,rust-book}.
When Quicksand ignores all data race candidates,
its results are competitive with SSS-MC-DPOR, although SSS-MC-ICB outperforms it slightly.
This is unsurprising: the seed subsets of preemption points
that QS-Sync-Only is limited to
(\cref{sec:quicksand-initial-pps})
are much less flexible than ICB's preemption strategy.
This result suggests that in future work,
Quicksand should consider using ICB in parallel with its default configuration when it finds no data race candidates to test.
I discuss this possibility further in \cref{sec:warpzone-heuristics}.

@@

Ultimately, using Limited Happens-Before for finding data race candidates allowed Quicksand to find more bugs,
while Pure Happens-Before allowed for reaching full verification faster.
I attribute this trade-off to the fact that
Limited Happens-Before need not wait to test many alternate thread interleavings before
finding a data race candidate to begin with;
%a potential data race candidate is confirmed;
rather, it can add new jobs to start testing potential races immediately.%
\footnote{Upcoming, \Cref{tab:drstatistix} will corroborate this conclusion:
the difference is most dramatic in {\tt alarm-\allowbreak{}simultaneous},
the test where Quicksand struggled most to finish even small subset jobs.}
%
On the other hand, Limited Happens-Before can get overwhelmed by too many false positives,
needing to refute such candidates by testing new state spaces for each one,
while Pure Happens-Before can refute false positives {\em en passant}
by testing alternate interleavings in its original state spaces.
%In both cases, the performance differences seem to become significant only after about 10 minutes of CPU time,
%which suggests... I don't actually know!
This suggests that
%each approach has merit, and that
model checkers which incorporate data race analysis should implement both modes
and offer the user to choose based on their desired and/or expected testing outcome.

@@

Beyond finding new bugs and completing full verifications with data-race preemption points,
I evaluated Quicksand's performance for classifying data race candidates in two ways:
its ability to check nondeterministic data races not reachable under a single-pass analysis (\cref{sec:quicksand-pps})
and its ability to suppress reallocation false positives (\cref{sec:quicksand-id-realloc}).
\Cref{tab:drstatistix} presents the results for this section.

@@

Some memory accesses may be hidden in a control flow path that requires a nondeterministic preemption to be executed
(\cref{sec:quicksand-pps}).
In such cases, a single-pass dynamic data race detector
might not achieve the coverage necessary
to identify a racing access pair as a candidate at all,
let alone check the resulting behaviour with such as Landslide.
I instrumented Landslide to report these to Quicksand
and counted how many such led to Quicksand finding new bugs when used as preemption points.
Such bugs could be considered {\em false negatives} of the single-pass approach.
The left half of \Cref{tab:drstatistix}
breaks down the types of bugs
found in each test case,
showing both the total number of data-race bugs
and the number among those that required such nondeterministic data races to expose.
% I denote these data race reports as {\em nondeterministic},
% while those that could be found on the first interleaving .....
%
To ensure a fair comparison, I disabled Quicksand's reallocation false positive suppression
(\cref{sec:quicksand-id-realloc}, itself evaluated in the next section)
for this experiment.
This prevents Landslide from suppressing an observed reallocation data race candidate on the first interleaving,
which would falsely classify it as nondeterministic,
even though a single-pass would not (indeed, should not) suppress such candidates.

\Cref{fig:dr-falsenegs}
%compares the types of data race candidates necessary to expose each data-race bug in the test suite.
visualizes the difference between single-pass and model-checking-enabled data race analysis.
The first and third series represent the bugs found using preemption points from single-pass data race candidates only,
% not entirely true, as portend could be given data race traces from an MC
%but they don't do it in their paper, so i feel comfortable making this claim
i.e., the state-of-the-art approach used by RaceFuzzer \cite{racefuzzer} and Portend \cite{portend}.
The second and fourth series show all data-race bugs Quicksand found,
which includes the former type as well as new bugs involving nondeterministic races.
QS-Limited-HB found a nice 69 data-race bugs in total, % nice
15 of which %could not be found with single-pass data race candidates alone.
required nondeterministic data-race preemption points to expose.
QS-Pure-HB is even more dependent thereupon, %on nondeterministic data-race preemption points,
%requiring nondeterministic data-race preemption points
requiring them in 21 cases among its 44 total data-race bugs.
Moreover, although the frequency
of these nondeterministic races varies across the different test cases
(for example, almost all in {\tt broadcast\_test} were nondeterministic; almost none in {\tt mutex\_test}),
they are still at least present in all tests,
meaning it is not just an issue of writing ``better'' test cases to avoid them.

@@

Note that I am not comparing how much testing time is required before identifying the data race candidates involved in each bug.
Single-pass data races can all be found after a single program execution,
while Quicksand may potentially take up to all 10 CPU-hours before identifying a nondeterministic data race.
However, prior work data race tools \cite{tsan,fasttrack},
being not integrated with a model checker,
are not intended to discover new candidates under subsequent runs.
Running a single-pass data race tool repeatedly for 10 CPU-hours could potentially uncover some nondeterministic candidates,
but stress testing's comparative problem with achieving reliable coverage is already well-understood
\cite{chess-icb,gambit},
so I hope the reader will consider this experiment enough evidence for model checking already.
Likewise, replay-based tools such as RaceFuzzer \cite{racefuzzer} and Portend \cite{portend}
depend upon the data race detector to provide an execution trace leading to each candidate.
This result suggests that
such tools could benefit from a similar feedback loop as is used in Iterative Deepening,
for example, to discover new transitively-reachable data races while testing initial ones,
even if full verification not necessarily be their goal.

@@

In \cref{sec:quicksand-realloc} I showed the soundness of
suppressing data race reports between two heap accesses when the surrounding memory was re-allocated in between.
\Cref{tab:drstatistix}'s ``Realloc FPs'' column shows the total number of such data race candidates for each test program,
totaling 965 across all tests.
% NB. Calculated this by grepping QS-LHB id-logs for "for realsies" (see quicksand pp.c).
Among these, only 64 were observed to avoid the reallocation in an alternate interleaving,
thereupon being promoted to real data-race preemption points.
\cref{sec:quicksand-realloc}'s proof guarantees the safety of pruning all state spaces resulting from the 901 others.
%
Among the 64 true data races, %which initially fit the malloc-recycle pattern,
none exposed a new bug when used as a preemption point.
This suggests that for other data race tools,
suppressing reallocation candidates may be a productive heuristic,
even if unsound without Iterative Deepening.
However, Quicksand was able to correctly identify the 64 violations of that heuristic (among 26 distinct tests),
and fall back to classifying them with DPOR.
